Practical Optional Types for Clojure

### 2015-02-27 metatheory.tex
theorem
NPE theorem
fix appendix vs. main labelling
formatting, remove/inline figures
finish overview
formatting examples
overview

\section{Metatheory}

Our proof for type soundness is similar to~\cite{TF10}. We add
errors and prove well-typed programs can never go wrong.

@@

{\javaassumptionsall}

For readability we define logical truth in Clojure.

{\istruefalsedefinitions}

We use an extra lemma to support our main soundness lemma. Consistency
ensures that occurrence typing does not claim the outer 
\mintinline{clojure}{x} is a keyword.

@@

{\consistentwithdefinition}

@@

\begin{lemma}\label{lemma:soundness}

  {\soundnesslemmahypothesis}
  \begin{proof}
    By induction on the derivation of the typing judgement.
  \end{proof}
\end{lemma}

% Uses above lemma:soundness label ^^^
{\soundnesstheorem}

{\wrongtheorem}

>>>>

\section{Metatheory}
\label{sec:metatheory}

Our proof for type soundness is similar to~\citet{TF10}. We add
errors and \wrong{} value and prove well-typed programs
rule out wrong programs.

@@

{\javanewassumption{main}}

%For readability we define logical truth in Clojure.

%{\istruefalsedefinitions{main}}

We use an extra lemma to support our main soundness lemma. Consistency
ensures that occurrence typing does not claim the outer 
\x{} is a keyword.

@@

{\consistentwithonlydef{main}}

@@

\begin{lemma}\label{main:lemma:soundness}

  {\soundnesslemmahypothesis}
  \begin{proof}
    By induction on the derivation of the typing judgement. 
    (Full proof given as lemma~\ref{appendix:lemma:soundness}).
  \end{proof}
\end{lemma}

We can now state our soundness theorem.

{\soundnesstheoremnoproof{main}}

{\wrongtheoremnoproof{main}}

Null-pointer exceptions are ruled-out in typed code.

{\nilinvoketheoremnoproof{main}}

### 2015-02-27 overview.tex
overview

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are pointer equal.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

Symbols support \emph{metadata}, an arbitrary map of
data. The syntax \clj{^File x}, shorthand for \clj{^{:tag File} x}, is a single expression that is the symbol \clj{x} 
with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File x] ...)}
the compile-time metadata is preserved in occurrences of \clj{x}
and utilized by the compiler.

\emph{Commas} are always whitespace.

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler using \clj{:require} to declare a
runtime dependency on Typed Clojure's core namespace
\clj{clojure.core.typed}.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.

Java methods and fields are called in prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
The canonical form for methods is \clj{(. target (method args*))} 
and for fields \clj{(. target field)} using the dot special form, which we utilise occasionally
for clarity.
Constructors are the class name with a dot suffix and arguments like \clj{(class. args*)}
or equivalently \clj{(new class args*)} with the \clj{new} special form.

The \clj{if} special form is a standard two-armed conditional.
The \clj{do} special form sequentially evaluates any number of expressions
left-to-right, returning the last expression or \nil{}.
The \clj{when} macro is a one-armed \clj{if} with an implicit \clj{do}
around arguments after the test, and returns \clj{nil} if the test fails.
The \clj{throw} special form is exactly Java's \java{throw}, which takes an
instance of \clj{Throwable} and throws it as an exception.
The \clj{fn} macro creates a function---it takes a parameter vector of symbols indicating parameter names
and any number of expressions in an implicit \clj{do} for the body of the function.
The \clj{let} macro takes a vector of bindings, a flat vector of binding-value pairs,
and scopes them in the body, an implicit \clj{do}.
Scoping works like \clj{let*} in Scheme---that is values in the binding vector can refer
to variables previously introduced in the same binder.
The typed
variant of \clj{fn} supports an optional keyword \clj{:-} and type following
each parameter to declare their static types, and similarly after
the parameter vector to declare the return type.

\emph{Map destructuring} is binding position syntax for pattern matching.
Parameters are replaced with a map containing entries of symbols to keywords that bind the symbols
to lookups on that keyword with the current argument---the \clj{:as} option
can follow which binds a symbol to the current argument. For example
$$
\clj{(fn [{^File x :x, :as m}] ...)}
$$ 
expands to a \clj{let} that binds \clj{m} to the argument and 
\clj{^File x} to \clj{(:x m)}.

In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
maps with known keyword entries with values of the associated type
The keys are not quoted because only keyword keys
are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
any number of types.

The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.

@@

This function takes a value that is either \emph{\#f} % mintinline really hates #
or a number, represented by an untagged \emph{union type}.
The `then' branch has an implicit invariant
that \rkt{x} is a number, which is automatically inferred with occurrence typing
and type checked without further annotations.

@@

@@

\inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}

@@

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path "dir/a") ;=> IllegalArgumentException ...
\end{minted}

We can use {\clj{defmethod}} to add a method to \clj{path}.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
Now \clj{path}'s dispatch table contains a method mapping
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. The call \clj{(path "dir/a")}
now dispatches to the method because
\clj{(isa? (class "dir/a") String)}
is true, and finally returns \clj{((fn [x] x) "dir/a")}.
\clj{isa?} is a subclassing check when provided with classes,
otherwise it is an equality check.
\begin{minted}{clojure}
  (isa? Keyword Object) ;=> true
  (isa? :a :a) ;=> true
\end{minted}

\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}

Typed Clojure does not statically ensure multimethod calls dispatch successfully---\clj{(path :a)} 
type checks but throws a runtime error.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Keywords are functions that look themselves up in their argument, so \clj{:op}
is our dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=25]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning
and use of HMaps.

@@

\clj{PayLoad} is either a HMap containing a file
or a string. We dispatch on \clj{:p} to distinguish the two cases---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.

>>>>

We now review the basics of the Clojure subset we use in the paper.

\paragraph{Syntax} Parentheses indicate \emph{applications}, brackets
delimit
\emph{vectors}, braces
delimit
\emph{hash-maps}
and double quotes delimit Java \emph{strings}.
\clj{nil} is exactly Java's \java{null}.
\emph{Symbols} begin with an alphabetic character.
A colon prefixed symbol like \clj{:a} is a \emph{keyword}.
\emph{Commas} are always whitespace.

\paragraph{Core forms} \clj{if} is a standard two-armed conditional.
\clj{do} sequences two expressions returning the latter.
\clj{throw} corresponds to Java's \java{throw}.
\clj{fn} creates a function from a parameter vector and a body expression.
\clj{let} takes a flat binding vector (which can refer to previous bindings like Scheme's \mintinline{scheme}{let*})
and a body expression.
\clj{def} binds a top-level \emph{var}.
\clj{defn} combines \clj{def} and \clj{fn}.

\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
associated with the key in the map, or \clj{nil} if there is none.
Keywords are functions that apply \clj{get} to their argument
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers;
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if both arguments are equal.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
\clj{nil} and \clj{false} are the false values.

\paragraph{Namespaces} \clj{ns}
declares a
\emph{namespace}
which manage var and class imports. 
\clj{:require} takes a number of namespace dependencies and \clj{:refer}
brings vars into scope by name.
\clj{:refer-clojure} \clj{:exclude} unmaps core vars from the current namespace.

\paragraph{Java Interoperability} Java methods and fields have prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.
Constructors are called with the class name with a dot suffix and arguments like \clj{(class. args*)}.

\paragraph{Metadata} Symbols support \emph{metadata}---the syntax \clj{^File x}
is a single expression of a symbol \clj{x} with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File x] ...)}
syntactic occurrences preserve metadata.

\paragraph{Destructuring} \emph{Map destructuring} is binding position syntax for pattern matching.
Parameters are replaced with maps of symbols to keywords that bind the symbols
to lookups on that keyword, optionally terminated by \clj{:as}
which aliases the parameter. For example
\clj{(fn [{^File x :x, :as m}] ...)}
expands to a \clj{let} binding \clj{m} to the first argument and 
\clj{^File x} to \clj{(:x m)}.

\paragraph{Static type syntax} Single quoted keywords are singleton types containing just the keyword.
Single quoted maps mapping keywords to types 
are HMap types representing
maps with known keyword entries with values of the associated type.
\clj{U} creates an \emph{untagged union} type, which can take
any number of types.

\paragraph{Typed Clojure Macros}
The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
The typed \clj{fn} has optional annotations by following a parameter
or binding vector by \clj{:-} and a type.

@@

This function takes a value that is either \emph{\#f} % mintinline really hates #
or a number, represented by an \emph{untagged} union type.
The `then' branch has an implicit invariant
that \rkt{x} is a number, which is automatically inferred with occurrence typing
and type checked without further annotations.

@@

This is a regular Clojure file compiled with
the Clojure compiler. It has a
runtime dependency 
on
\clj{clojure.core.typed},
Typed Clojure's core namespace.
Programmers import typed versions of `typed' macros and qualified types to ease annotation.
The \clj{check-ns} function type checks the current namespace.

@@

\begin{exmp}
\inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}

We can use {\clj{defmethod}} to add a method to \clj{path}.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{File} to the function
\clj{(fn [^File x] (.getPath x))}. 
The call 
$$
\clj{(path (File. "dir/a"))}
$$
now dispatches to the \clj{File} method because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [^File x] (.getPath x)) "dir/a")}.
$$

The \clj{isa?} function first tries an equality check
on its arguments, then if that fails
and both arguments are classes a subclassing
check is returned.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? Keyword Object) ;=> true
\end{minted}

In the full definition for \clj{path} we add a method for \clj{nil}
which uses the equality check provied by \clj{isa?}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=11]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}

Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
type checks because \clj{:a} agrees with the parameter type \clj{Any},
but throws an error at runtime.

The flexibility of \clj{isa?} is key to the generality of multimethods. 
We can dispatch on the \clj{:op} key 
of our HMap AST \clj{Expr}.
Since keywords are functions that look themselves up in their argument, we simply
use \clj{:op} as the dispatch function.

\begin{exmp}
\inputminted[firstline=5,lastline=31]{clojure}{code/demo/src/demo/eg5.clj}
\end{exmp}

@@

The final example combines everything we will cover for the rest of the paper:
multimethod dispatch, reflection resolution via type hints, Java method
and constructor calls, conditional and exceptional flow reasoning,
and HMaps.

@@

We dispatch on \clj{:p} to distinguish the two cases of \clj{FSM}---for example on \clj{:F}
we know the \clj{:file} is a file.
The body of the first method uses type hints to resolve reflection
and conditional control flow to prove null-pointer exceptions are impossible.
The second method is similar except it uses exceptional control flow.
